```{r load libraries}
#| warning: False
library(tidyverse)
```

```{r countries}
countries <-
  who |> 
  distinct(country) |> 
  format_delim(delim="\\n",eol = "|", col_names = FALSE)
```

```{r}
rawdons <-rawdons |> 
    mutate(
      Country = sapply(str_extract_all(Description, str_c("United Kingdom|Côte d'Ivoire|Hong Kong",countries)), function(x) paste(x, collapse=','))) |>
    mutate(Country=str_match(Country,pattern = r"([\w+\s+\w+]+)"))
```
After adding the country
add the isocodes by merging
```{r}
isocodesaug <- isocodes |> 
  select(iso2,iso3,Country)

rawdons <-rawdons |> 
  mutate(Country = str_replace_all(Country,
    c("Democratic Republic of the Congo"="Congo Democratic Republic of the", 
      "Republic of Korea"="Korea Republic of",
      "United Republic of Tanzania"="Tanzania United Republic of"))) |> 
  left_join(isocodesaug, join_by(Country))
```

add icd based on disease column
add disease column
```{r}
uniqdisease <- uniquedons |> 
  select(DONs, Disease) |> 
   separate_longer_delim(DONs, delim = ",",) |> 
  mutate(DONs=str_remove(DONs," "))

```
join the diesease and dons 
```{r}
rawdons <-rawdons |>
  left_join(uniqdisease, join_by(ID==DONs)) |> 
  mutate(Disease = ifelse(is.na(Disease), ID, Disease))
```
Join the icd codes
```{r}
rawdons <-rawdons |>
  left_join(icd1011, join_by(Disease))
```

Delete DONs reporting other types of events, such as meetings of international experts (see DON0133), recommendations on treatments (DON0153), or travel requirements (DON0107 and DON0198). 
```{r}
rawdons <- rawdons |> 
  filter(ID!="DON0133"& ID!="DON0153"& ID!="DON0107"& ID!="DON0198") |> 
  arrange(Country)
```
Remove duplicates by deleting all the DONs registering the information on the same disease for the same country more than once in a given calendar year 
```{r}
rawdons <- rawdons|> 
  mutate(Date = str_match(Date, pattern =r"(\d+)"))
```
Chnage what is happenign in 2019
```{r}

rawdons <- rawdons |> 
  mutate(Country = ifelse((Disease=="Anthrax, unspecified" & Date=="2001"), "United States", Country )) 
 
```

```{r}
 rawdons |> 
  # distinct(Date, Country,Disease) 1120
  filter(is.na(Country))
  # summarise_all(~sum(is.na(.)))
```

